\hspace{1cm} Do syntactic and semantic similarity lead to interference effects? \newline Evidence from self-paced reading and event-related potentials using German

### 2023-12-18 paper/main.tex
added SPR posterior predictions

\textcite{mertzen} presented simulations that show approximately equally sized interference effects 
for both the syntactic and semantic manipulations. Their model assumed that three cues were used at the verb: \{$\pm$ grammatical subject\}, \{$\pm$ same clause\} and \{$\pm$ animacy\}. The upper parts of Figure \ref{fig:model_predictions} a and b reproduce the \textcite{mertzen} simulations. \ref{fig:model_predictions_rts} shows the prior predictive distributions of by-condition retrieval times that were generated by sampling from a prior distribution ($Beta(a=4,b=6)$) for the free scaling parameter F and then generating data from the model while holding all other parameters in the model at fixed values. The F parameter scales the activation to a retrieval time. The prior $Beta(a=4,b=6)$ was chosen in earlier work \parencite{jaeger_etal_2020}. Figure \ref{fig:model_predictions_effects} shows the effects of interest computed based on the condition means shown in \ref{fig:model_predictions_rts}.

@@

Could the absence of syntactic interference be reconciled with a cue-based retrieval account? We speculated that one possibility might be that the parser tracks hierarchical structure and that this hierarchical information is used when searching for a noun in memory. Specifically, the relevant syntactic cues that might be used at the verb are not  \{$\pm$ grammatical subject\} and \{$\pm$ same clause\} but rather a composite cue \{$\pm$ subject-in-same-clause\}. In other words, in the spirit of the  \textcite{dillon2013} and \textcite{Sturt2003} proposal, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval, we evaluated the predictions of the model assuming that the cues are \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. The results of the prior predictive simulations are shown in 
the lower parts of Figure \ref{fig:model_predictions} a and b. This model now shows the observed pattern: a semantic interference effect but no syntactic interference. One limitation here is that the prior predictive distribution of our post-hoc  model still overestimates the semantic interference effect (95\,\% CrI [10, 67]\,ms; the largest effect observed the SPR study was [4, 12] ms at the critical region). This overestimation has to do with the fact that the prior on the F parameter is relatively spread out; it would be  possible to fit the parameter to the SPR data.

>>>>

\textcite{mertzen} presented simulations that show approximately equally sized interference effects 
for both the syntactic and semantic manipulations. Their model assumed that three cues were used at the verb: \{$\pm$ grammatical subject\}, \{$\pm$ same clause\} and \{$\pm$ animacy\}. The upper parts of Figure \ref{fig:model_predictions} a and b reproduce the \textcite{mertzen} simulations. \ref{fig:model_predictions_rts} shows the prior predictive distributions of by-condition retrieval times that were generated by sampling from a prior distribution ($Beta(a=4,b=6)$) for the free scaling parameter \textit{F} and then generating data from the model while holding all other parameters in the model at fixed values. The \textit{F} parameter scales the activation to a retrieval time. The prior $Beta(a=4,b=6)$ was chosen in earlier work \parencite{jaeger_etal_2020}. Figure \ref{fig:model_predictions_effects} shows the effects of interest computed based on the condition means shown in \ref{fig:model_predictions_rts}.

@@

Could the absence of syntactic interference be reconciled with a cue-based retrieval account? We speculated that one possibility might be that the parser tracks hierarchical structure and that this hierarchical information is used when searching for a noun in memory. Specifically, the relevant syntactic cues that might be used at the verb are not  \{$\pm$ grammatical subject\} and \{$\pm$ same clause\} but rather a composite cue \{$\pm$ subject-in-same-clause\}. In other words, in the spirit of the  \textcite{dillon2013} and \textcite{Sturt2003} proposal, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval, we evaluated the predictions of the model assuming that the cues are \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. The results of the prior predictive simulations are shown in the lower parts of Figure \ref{fig:model_predictions} a and b. This model now shows the observed pattern: a semantic interference effect but no syntactic interference. One limitation here is that the prior predictive distribution of our post-hoc  model still overestimates the semantic interference effect (95\,\% CrI [10, 67]\,ms; the largest effect observed the SPR study was [4, 12] ms at the critical region). This overestimation has to do with the fact that the prior on the \textit{F} parameter is relatively spread out.

To reduce the overestimation of the effect size of the semantic interference effect, we generated posterior predictive distributions for the effects of interest by fitting the \textit{F} parameter to the SPR data. New data was simulated using only the values of F which produced a semantic interference effect in the range of the semantic interference effect in our SPR data (mean $\pm \,2\times$SE). This applied to only 355 out of 20,000 sampled values of \textit{F}. We fitted \textit{F} only to the semantic interference effect (and not the syntactic interference effect or interaction) to avoid overfitting. Figure \ref{fig:model_predictions_posterior} shows the posterior predictive distributions of the effects of interest in comparison to the prior predictive distribution and our SPR data. The posterior predictions match the syntactic and semantic interference effects in our data, but underestimate the interaction.

\begin{figure}[h]
    \caption{Prior and posterior predicted effect sizes, i.e., retrieval time differences, for syntactic and semantic interference and their interaction using the two cues \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. The data with 95 \% credible intervals is shown in red.}
    \label{fig:model_predictions_posterior}
    \centering
    \includegraphics[width=\textwidth]{paper/images/priorpred_vs_postpred_vs_data.png}
\end{figure}
